---
title: "What is the role of superenhancers?"
author: "Kevin Stachelek"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: 
    number_sections: TRUE
  pdf_document:
    fig_caption: yes
    latex_engine: xelatex
bibliography: superenhancers.bib
csl: "/home/ks/.cache/R/seasl/styles/cell.csl"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r , include=FALSE}
library(seasl)
library(stchlk.notes)
pwd <- getwd()
stchlk.notes::loadkey(pwd)

```


# Enhancers 

## What is the importance of enhancers in gene expression regulation?

* Enhancers regulate spatiotemporal gene expression. Enhancers also require transcription factor binding and histone modification but how much is unclear [@Barakat2018]

## What methods exist to study enhancer function?

* @Barakat2018 combined Chip with parallel reporter assay to id functional enhancers in ESCs. (Chip-STARR-Seq)

## What is the role of enhancers in ESC development? 

* Only a minority of NANOG, OCT4, H3K27ac, and H3K4me1 bound regions function as enhancers [@Barakat2018]

* There is a new group of functional enhancers that are active in ESCs but are associated with generic cell processes (The extended enhancer module).

* The extended enhancer module is characterized by reduced binding of pluripotency associated TFs and histone marks. 

* @Barakat2018 identified an enhancer set associated with non-ESC-specific processes.

## Can we predict enhancers? 

* TF binding is linked with enhancer activity but no individual TF, histone mark or combination could unequivocally predict enhancer activity [@Barakat2018]. 

## How do transposable elements relate to enhancers?

* Transposable elements may associate with enhancers, but only some exhibit activity [@Barakat2018]. 

* Within super-enhancers large tracts are nonfunctional, restricted to small subdomains [@Barakat2018]. 

## Relationship betwene gene expression and promoter-enhancer interaction

* Observed correlations between gene expression, promoter-enhancer interactions and enhancer RNAs [@Sanyal2012]. 

## What is the kB scale of promoter-enhancer interactions? 

* Long-range interactions have a bias for elements located ~120 kB upstream of the TSS [@Sanyal2012].

* Long-range interactions are often not blocked by sites bound to CTCF and cohesin. So many of these sites do not demarcate physically insulated gene domains [@Sanyal2012]. 

* only 7% of looping interactions involve the nearest gene. Genomic proximity is not a straightforward predictor for long-range interactions [@Sanyal2012]. 

# Superenhancers 

## Definition of super-enhancers 

* super-enhancers are dense TF platforms enriched for active chromatin marks[@Shin2016] 

* Super-enhancers are large clusters of transcriptional enhancers that drive expression of genes that define cell identity [@Hnisz2013]. 

* Many genes that determine cell identity are regulated by clusters of mediator-bound enhancer (superenhancers) [@Hay2016]

* Large linear tracts of chromatin, super-enhancers (SEs) have been identified that function to regulate lineage-specific gene expression [@Whyte2013]

* Compared to traditional enhancers, SEs have increased binding of Mediator, specific histone marks, and lineage-specific TFs [@Barakat2018]. 


## Super-enhancers vary on a cell-type specific basis 


* Improved understanding of super-enhancers could be had by identifying superenhancer elements across cell types. 

* Super enhancers associate with genes that control and define cell types [@Hnisz2013]. 

* Disease-associated variation is enriched in super-enhancers of disease-relevant cell types [@Hnisz2013]. 

* Cancer cells generate super-enhancers at oncogenes and other drivers [@Hnisz2013]. 

## Model of Super-enhancer function

### Super-enhancer constituents act additively and independently

* @Hay2016 Generated a series of mouse models. Deleted each of the five regulatory elements of the a-globin super-enhancer individually and in combination. 

* They found that each constituent enhancer of the a-globin super-enhancer acts independently and additively [@Hay2016].

* a-globin super-enhancer impacts hematological phenotype, gene expression, chromatin structure, and chromosome conformation [@Hay2016]. 

### Chip-STARR-Seq dissects super-enhancers into small functional units

* SEs were identified by H3K27Ac enrichment. And alignment of ChIP-STARR-seq data to SEs showed correlation between RPP and H3K27Ac intensity [@Barakat2018]. 

* By modifying regions of the FGFR1 SE and observing effects via ChIP-STARR-seq it was observed that   The FGFR1 SE is composed of small units with enhancer activity [@Barakat2018]. 

* After looking at global patterns of reads per plasmid (RPP) it was found that enhancers contain only a small percentage of active plasmids in their bounds. The fraction is only slightly higher in SEs [@Barakat2018]. 

* It appears that enhancer activity is limited to small domains within the SE that frequently overlap with TF binding sites. 

* The observed chromatin signatures at SEs, such as H3K27Ac, might be a consequence of enhancer activity from much smaller units. 

* SEs constituents might function:
  + alternatively as independent and additive enhancers [@Hay2016, @Moorthy2017]
  + in a temporal and functional hierarchy [@Shin2016]
  + as interdependent units [@Hnisz2013]
  
* Caveat: Chip-STARR-Seq does not fully recapitulate the endogenous chromatin context.

* "High-throughput sequencing studies have attempted to predict ESC enhancers on the basis of histone marks, TF binding, or DNaseI hypersensitivity (Hawkins et al., 2011; Rada-Iglesias et al., 2011; Xie et al., 2013). However, the overlap between enhancers predicted from these studies is limited [@Barakat2018]"

## Superenhancers targets for Cancer Therapy

* Chromatin regulators are targets for cancer therapy [@Loven2013]

* Why does inhibition of chromatin regulators have gene-specific effects in tumor cells?

* How can inhibition of BRD4 lead to selective inhibition of myc in multiple myeloma (MM)?

* BRD4 and mediator co-occupy 1e3 enhancers associated with active genes [@Loven2013]. 

* BRD4 and Mediator co-occupy a few very large super-enhancers associated with MYC and other genes [@Loven2013].

* Treatment with JQ1 (BED-bromodomain inhibitor) leads to loss of BRD4 at super-enhancers and transcription elongation defects at super-enhancer genes [@Loven2013]. 

## What is the role of Superenhancers in ESC development?

* TFs Oct4, Sox2, Nanog bind enhancers and recruit Mediator to activate gene expression in ESCs [@Whyte2013]

* ESC master TFs form enhancer domains at most genes that control pluripotent state [@Whyte2013]. 

* reduced levels of Oct4 or mediator cause loss of super-enhancers relative to other genes [@Whyte2013].

* in differentiated cells superenhancers containing cell-type-specific master TFs are found at genes that define cell identity [@Whyte2013]. 

## Mechanism of regulation of super-enhancers

### Cohesin

#### Role of Cohesin

* Cohesin is a protein which mediates sister chromatid cohesion during the cell cycle. Cohesin also forms a complex with CTCF and active gene regulatory elements to form long-range interactions between binding sites [@Ing-Simmons2015]. Chromosome conformation capture shows that cohesin's main role in interphase is mediating interactions with chromosome compartments [Ing-Simmons2015]. 

#### How do cohesin-mediated interactions contribute to regulation of gene expression?

* Ctcf and cohesin are enriched at enhancers, and especially at enhancer arrays, also called super-enhancers [@Ing-Simmons2015]. Global and local chromosome conformation capture experiments have shown that enhancer elements associate both linearly and in three dimensions [@Ing-Simmons2015]. 

#### Is spatial enhancer clustering facilitated by cohesin? 

* Conditional deletion of cohesin from cells preserves enhancer position and transcription but weakens interactions between enhancers and promoters [@Ing-Simmons2015]. Half of all deregulated genes following cohesin deletion lie in the vicinity of enhancer elements, so it is likely that cohesin regulates gene expression through spatial clustering. 

* It is possible to create a model of cohesin dependent gene regulation where spatial clustering of enhancer elements is the unified mechanism for enhancer-promoter connection and insulation [@Ing-Simmons2015]



# Comparing Enhancers and Superenhancers

## Structural differences enhancers v. superenhancers?

* Super-enhancers consist of clusters of enhancers that are occupied by master regulators and Mediator [@Whyte2013].

* Super-enhancers differ from enhancers by:
  + size
  + TF density
  + ability to activate transcription
  + sensitivity to perturbation [@Whyte2013]


## Functional differences enhancers v. superenhancers? 

### Enhancers and super-enhancers have equivalent regulatory roles in escs by regulation of single or multiple genes

* Trans enhancers maintain cell-type-specific expression. They drive cell fate changes during development [@Moorthy2017]. 

* Highly expressed genes are associated with individual enhancers, ex. locus control regions [@Moorthy2017].

* it is predicted that stretch-enhancers (super-enhancers) regulate critical cell-identity genes. 

### Crispr/cas-9 deletion to compare enhancers and superenhancers. 

* The effect of deleting enhancers is variable, results in gene expression changes from 12-92% [@Moorthy2017]. 

* partial deletion of enhancers that removed only one enhancers a sub cluster shows redundant control of gene by multiple enhancers in a super-enhancer.

* highly expressed genes in ESCs are not necessarily associated with a super-enhancer [@Moorthy2017].

* super-enhancer prediction ignore 81% of active regulatory elements predicted by co-binding of pluripotency-associated TFs. 
  + but these are very critical. 
  
* sometimes super-enhancers regulate clusters of paralogous genes (significance?) [@Moorthy2017].


### Superenhancers in Mammary Tissue Development

* What is the role of super-enhancers in the mammary gland, which is dynamically regulated during pregnancy? 

* chip-seq for STAT5A, glucocorticoid receptor, H3K27Ac, and MED1 identified 440 Mammary-specific super-enhancers [@Shin2016]. 

* Half of mammary-specific super-enhancers are associated with genes activated during pregnancy [@Shin2016].

* Generated mice carrying mutations in STAT5 superenhancer binding sites. The most distal site had the greatest enhancer activity [@Shin2016]. 

* mutation analysis showed that 1e3 fold induction during pregnancy relied on all enhancers [@Shin2016]. 
* disabling the binding sites of STAT5, NFIB, ELF5 in the proximal enhancer incapacitated the super enhancer [@Shin2016]. 

* There is a temporal and functional enhancer hierarchy [@Shin2016]. 

* Mammary-specific super-enhancers demonstrate cell-type specific expression of hormone-sensing genes [@Shin2016]. 


# TADs

## Definition of TADs

* interphase chromosomes occupy distinct space. There are many models of chromosome folding [@Cremer2010]. 

* @Dixon2012 investigated the three dimensional organization of human and mouse ESCs and terminally differentiated cells using HI-C technology.

* Megabase-sized chromatin interaction domains are termed 'topological domains' [@Dixon2012]

* topological domains correlate with sites of heterochromatin constraint. And these domains are stable across cell types and conserved across species [@Dixon2012]. 

* Chromosome conformation capture show that interphase chromatin is partitioned into Mb-sized compartments and sub-Mb-sized topological domains [@Seitan2013]. 

* Compartmentalization facilitates gene and reg. element matching [@Seitan2013].

## How are TADs demarcated? 

* Boundaries of topological domains are enriched for the insulator CTCF, housekeeping genes, tRNAs and SINE retrotransposons [@Dixon2012]. 

* Non-promoter associated marks, such as H3K4me1 (associated with enhancers) are not enriched or are specifically depleted at boundary regions [@Dixon2012]. 

## How are TADs formed?

* Cohesin controls chromosome topology to enable DNA repair and chromosome segregation in cycling cells [@Seitan2013]. 

* Cohesin also associates with active enhancers and promoters and with CTCF to form long-range interactions for gene regulation [@Seitan2013]. 

* Architectural compartments are maintained in non-cycling mouse thymocytes after genetic depletion of cohesin in vivo [@Seitan2013]. 

* Cohesin is required for specific long-rang interactions within compartments where cohesin-regulated genes reside [@Seitan2013]. 

* Cohesin depletion diminishes interactions between cohesin-bound sites. Cohesin depletion also makes alternative interactions between transcriptional activation and repression features more prominent--resulting in changes in gene expression [@Seitan2013]. 

* Cohesin-mediated long-range interactions facilitate discrete gene expression states within preexisting chromosomal compartments [@Seitan2013]. 

## Is there variation in chromatin conformation within TADs?

* @Giorgetti2014 developed a polymer model to extract the full repertoire of chromatin conformations within TADs from population-based 3C data. 


## How are TADs important in promoter-enhancer interactions?

* @Chen combined genome editing and multi-color live imaging to visualize physical enhancer-promoter interaction in single cells.

* Identify three topological states with different transcription kinetics. Sustained proximity of the enhancer to its target is required for activation [@Chen]. 

* transcription affects 3d topology by enhancing temporal stability of the proximal conformation and spatial compaction [@Chen]. 

* Long range activation results in transcriptional competition at the locus [@Chen]. 

## How does enhancer contact relate to TAD structure?

* The model developed by @Giorgetti2014 predicts physical distances and chromosomal contact variation among cells. It also identifies interactions within single TADs that stabilize boundaries between TADs and allows the identification and validation of structural elements within TADs [@Giorgetti2014]. 

* Combining polymer model with DNA FISH and RNA FISH in Xic, @Giorgetti2014 dissected the relationship between transcription and spatial proximity to cis-regulatory elements. 

* Contacts between potential regulatory elements occur in the context of fluctuating structures rather than stable loops and such fluctuations may contribute to asymmetric expression in Xic [@Giorgetti2014].

## How do TADS mediate enhancer-promoter interaction?

### enhancer-promoter proximity mediated by TADS may be necessary but not sufficient for expression

* Limb-specific Shh expression is regulated by the ZRS enhancer. @Williamson2016 analyzed the higher-order chromatin conformation of Shh in expressing and non-expressing tissues by FISH and 5C. 

* Light microscopy identified elevated frequencies of Shh/ZRS colocalization only in the Shh-expressing regions of the limb bud due to enhancer-promoter loop formation.

* in all tissues at all stages Shh-ZRS spatial distance are shorter than the distance to an intervening neural enhancer [@Williamson2016]. 

* 5C identified a TAD over the Shh/ZRS genomic region and enriched interactions between Shh and ZRS through E111.5 [@Williamson2016]. 

* Shh/ZRS colocalization correlates with spatiotemporal limb-bud specific Shh expression, but close Shh and ZRS proximity in the nucleus occurs regardless of whether the gene or enhancer is active [@Williamson2016]. So expression is not specifically regulated by enhancer-promoter proximity in this case. Rather, constrained chromatin configuration optimizes the opportunity for the active enhancer to locate and instigate the expression of Shh [@Williamson2016]. 

## How does interTAD arrangement impact expression of border genes?

* The transcriptional activation of Hoxd genes during vertebrate digit development involves modifications in 3D interactions within and around the HoxD gene cluster [@Fabre2015]. 

* @Fabre2015 used 3D DNA FISH to assess the spatial organization of chromatin at and around the HoxD gene cluster. Although the two TADs are tightly associated, they appear as spatially distinct units. 

* @Fabre2015 measured the relative position of genes within a cluster and find that they segregate over long distances, so a physical elongation of the HoxD cluster can occur. 

* @Fabre2015 used STORM to find that tissues with distinct transcriptional activity exhibit differing degrees of elongation. The spatial separation of Hoxd genes is strongest in those tissues where they are highly expressed.

* The morphological change of the HoxD cluster in developing digits is associated with its position at the boundary between two TADs [@Fabre2015]. 

* Variations in fine-scale architecture of gene clusters suggest causal links among spatial configuration, transcriptional activation, and flanking chromatin context [@Fabre2015]. 

* In the case of HoxD, all enhancers sharing a particular specificity are found within the same TAD [@Fabre2015].

* So a model is suggested wherein reorganization follows a global transition from one set of regulatory contacts to another--between two TADs located on either side of the HoxD locus [@Fabre2015]. 

## Long 

* "increased numbers of TF motifs at an enhancer should be positively correlated with increased nucleosomal eviction, DNA binding, and, consequently, elevated gene-expression output [@Long2016]."

* "at some developmental enhancers, there is evidence for step-wise licensing by lineage-determining TFs, also known as master regulators, or pioneer factors, which directly bind nucleosomal DNA to prime enhancers for activation [@Long2016]."

* "Thus, combinatorial deployment of multiple sub-optimal enhancers may help to promote specificity of regulation without sacrificing signal strength."

* "One hypothesis is that enhanceosomes may have potential roles in mediating switch-like transcriptional activation of genes requiring strict regulatory inputs from multiple signaling path- ways (Spitz and Furlong, 2012). Consistent with this theory, deeply conserved enhancers tend to fall as clusters near genes encoding developmentally important transcription factors"

## References