---
title: "accessibility_bell.Rmd"
author: "Kevin Stachelek"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r , include=FALSE}
library(stchlk.notes)
pwd <- getwd()
stchlk.notes::loadkey(pwd)

```

* histone modifications regulate gene expression by modulating DNA accessibility. 

* _Bell_ _et_ _al_ measured gDNA accessibility in _D._ _melanogaster_ by combining M.Sssl footprinting with meDNA immunoprecipitation. 

* me accessibility marks different distribution of active and repressive histone mods and sites of transcrip and replic init. 
* DNA access is increased at active promoters and chrom regions that are hyperAc at H4K16, esp. in male X chrom. 
  + so dosage compensation is facilitated by permissive chromatin structure
  
* inactive chromos domains with H3K27me3 are least accessible suggesting polycomb-mediated chromatin compaction. 

* higher accessibility at chrom that replicate early and at sites of replication init. 

overall: differential histone mod patterns and the organization of replication have effects of exposure of DNA template. 


