---
title: "enhancers_chen.Rmd"
author: "Kevin Stachelek"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r , include=FALSE}
library(stchlk.notes)
pwd <- getwd()
stchlk.notes::loadkey(pwd)
```

# intro
enhancers play a role in the control of gen expression during development.

they span a lenght of 50-1500 bp and are cis-regulatory. 

Enhancers stimulate transcription from core promoters by recruiting activators and repressors

There are between 20e4 and 1e6 enhancers in the human genome. 

a large fraction of promoters are located large distances from corresponding promoters

in D. melanogaster 30% occur over 20kb, with intervening genes

How do enhancers communicate with target promoters over large distances?

methods already developed:
FISH
chromosomal conformation capture

lack a dynamic characterization to distinguish transient contact from stable structures and establish causality

have developed a live-imaging method at single cell level

# results

placed a reporter gene 142kb from D. melanogaster __eve__ locus

__eve__ has five enhancers that drive striped pattern on the early fly. 

chromatin fiber displays random movements at this distance. creates hurdle for specific interactions. 

also included insulator _homie_ which supports stable enhancer-promoter loop by self-pairing with endogenous _homie_ near _eve_ 

they find reporter gene only in striped pattern areas --specific)

need to visualize:
+ location of endogenous _eve_ enhancers
+ promoter of hte reporter
+ transcriptional activity of reporter

design three-color system. 

use stem-loop labeling cassetes, [MS2](https://en.wikipedia.org/wiki/MS2_tagging) in the _eve_ gene, and [PP7](https://en.wikipedia.org/wiki/CRISPR-Display) were introduced into the reporter gene.

fluorescent coat proteins bind nascent stem-loops on transcription

fluorescence also marks the nuclear position of the _eve_ enhancers 

also used another system to label the position of the reporter gene independent of its activity. 

with this system they did three-color time -lapse confocal microscopy on a two-hour old embryo

red: transcription of reporter
blue: transcription of eve
green: position of reporter gene

control experiments:
## establish baseline
generate a construct where all three colors are within 2.0kb (establish baseline)

error accounts for 10% of variance in distance measurement. 

## genomic labeling introduce artifact?
move around genomic labeling construct and compare to endogenous and/or common system. 
no difference found


close connectio nbetween transcriptoin and physical proximity of e-p pair by eye

smaller distance between blue-green found when red is on (transcription) than when red is off (no transcription)

change in e-p (blue-green) distance can give mean squared displacement (msd) in red-on and red-off cases. e and p get closer and stay closer more in red-on than in red-off. 

## how important is e-p proximity over time?
switches between red-on and off occurred at points of high e-p proximity 



transcription initiation ceases after enhancers and promoters separate. though polII continues transcription already underway after separation. 

## three topological states

These three florescent foci thus provide the means to measure
the physical distance between the enhancers and the reporter, as
well as to monitor the reporter’s transcriptional activity.

1. generated a synthetic construct in which all three fluorescent proteins are co-localized within a genomic distance of 2.0 kb
2. test whether genomic labeling introdcues perturbations in the system
3. 


