---
title: "moorthy_enhancers.rmd"
author: "Kevin Stachelek"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r , include=FALSE}
library(stchlk.notes)
pwd <- getwd()
stchlk.notes::loadkey(pwd)

```

# Enhancers and super-enhancers have equivalent regulatory roles in escs by regulation of single or multiple genes

* Trans enhancers maintain cell-type-specific expression. They drive cell fate changes during development. 

* Highly expressed genes are associated with individual enhances, ex. locus control regions.

* it is predicted that stretch-enhancers (super-enhancers) regulate critical cell-identity genes. 

## Can use crispr/cas-9 deletion to study enhancers. 

* The effect of deleting enhancers is varialbe, results in gene expression changes from 12-92%. 

* partial deletion of enahncers that removed only one enhancers a subcluster shows redundant control of gene by multiple enhancers in a super-enhancer.

* highly expressed genes in escs are not necessarily associated wit ha super-enhancer.

* super-enhancer prediction ignore 81% of active regulatory eleemnts predicted by cobinding of pluripotency-associated tfs. 
  + but these are very critical. 
  
* sometimes super-enhancers regulate clusters of paralogous genes (significance?)

