---
title: "sharma_computational.Rmd"
author: "Kevin Stachelek"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r , include=FALSE}
library(stchlk.notes)
pwd <- getwd()
stchlk.notes::loadkey(pwd)

```

# Background 

* nuclear receptors are a class of tf that act with c-regulators and repressors to maintain: metabolism, development, and reproduction. 

* NR are activated through binding of small ligands, replaced by drug molecules, so promsiing drug targets. 

* trans regulation of genes that encode NRs is important for understanding function and disease relevance of NRs. 

* NR share evo history but have different expression patterns. can be ubiquitous or spatiotemporal complex. 

# Method 

* relationship between long-range reg of NR and function. id nr genes based on counts of non-coding elements. 

* validate results by rna-seq and ChIP-seq from ENCODE. 

* Find NR genes involved in dev roles have long-range mech of trans reg with distinct cis-reg contnt. Feature:
  + clusters of highly conserved non-coding elemetns in regions over Mb
  + long and mult cpg islands 
  + bivalent promoter and 
  + higher enrichment of enhancer mark. 


* Nr genes involved in tissue spec lack these. have simple trans controls and variety of mech for paralog prod. 

* combinatorial patterns of histone maps show reg landscape of gene familys. support long-range reg is inidicative of NR functional role compared to existing classifications. 
