---
title: "simmons_spatial.Rmd"
author: "Kevin Stachelek"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r , include=FALSE}
library(stchlk.notes)
pwd <- getwd()
stchlk.notes::loadkey(pwd)

```

# Abstract

* cohesin complex associates with CTCF and active gene reg elements to form long-range interaction between binding sites. 

* Also mediates sister chromatid cohesion during hte cell cycle. 

* chromosome conformation capture shos that cohesin's main role in interpahse is mediating interxn with chromoome compartments. 

* How do cohesin-mediated interactions contribute to regulation of gene expression

* ctcf and cohesin is enriched at enhancers esp at enhancer arrays (super-enhancers)

* global and local chrom conf capture shows that enhancer elements associate linear and 3d. 

* spatial enhancer clustering facilitate by cohesin. 

* conditional deleetion of cohesin from cells preserves enhancer position and transcritpion but weakens interactions between. 

* half of dereg genes in vicinity of enhancer elemetns. so cohesin regulates gene expression through spatial clustering. 

* model for cohesin dependent gene regulation where spatial clustering of enhancer elements is unified mech for e-p connection and insulation 