---
title: "TAD_franke.Rmd"
author: "Kevin Stachelek"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r , include=FALSE}
library(stchlk.notes)
pwd <- getwd()
stchlk.notes::loadkey(pwd)

```

* Chromosome conformation capture has id'd topologically associated domains (TADS)

* separated from each other by boundary regions. 

* restrict the contacts that enhancers establish with target genes

* show by capture Hi-C and 4C-seq that genomic duplication in patient cells and gmo mice results in formation of neo-TADs

* Duplications of non-coding DNA in _Sox9_ TAD cause f to m sex reversal. they show increased contact of duplicated regions in the TAD but no change in TAD structure. 

* Contrast: overlapping duplications that extend into a neighboring TAD (inter-TAD) resulted in the formation of a new chromatin domain (neo-TAD) isolated from the rest of the genome. 

* So there is an _insulation_ phenomenon that protects from inter-TAD duplications 

* However, incorporation of incorporation of a flanking gene, in the neo-TAD results in ectopic contacts of the flanking gnee with the duplicated part of the regulatory region, misexpression, and path. pheno. 

* conclusion: TADs are genomic regulatory units with internal stability that are altered by genomic variations. 

* has bearing on copy number interpretation. also relevant for evolution, since copy number alterations are important for evolution. 





